Determining the existance and direction of causality in the face of measurement error

### 2016-02-25 docs/manuscript.Rmd
updated

---
title: "Determining the direction of causality in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: true
bibliography: manuscript.bib
csl: nature.csl
---

@@

Here, however, arises a potential issue: often the biological effect of a SNP is not known. If there exists a putative association between two variables, with a SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. For example genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. Another situation is where the causal direction between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?

@@

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of mediation based methods. Mendelian randomisation is robust in the face of measurement error and, given that the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance to the trait of interest. But on many occasions this is not the case, and it may be tempting to resort to using methods based on mediation in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers two main advantages. First, it uses a formal statistical framework to test for the robostness of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is much less likely to infer the wrong direction of causality compared to CIT.

There are a number of limitations to the analysis that we performed. We focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.

>>>>

---
title: "Determining the existance and direction of causality in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  <!-- word_document -->
  pdf_document:
    keep_tex: true
bibliography: manuscript.bib
csl: nature.csl
---

@@

Here, however, arises a potential issue: often the biological effect of a SNP is not known. If there exists a putative association between two variables, with a SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. For example genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. To illustrate a trivial example - the same FTO variant that is famously associated with BMI is also associated with C-reactive protein (CRP). It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would infer that CRP causes BMI. Another situation is where the causal direction between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?

@@

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of mediation based methods. Mendelian randomisation is robust in the face of measurement error and, given that the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance to the trait of interest. But on many occasions this is not the case, and it may be tempting to resort to using methods based on mediation in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers two main advantages. First, it uses a formal statistical framework to test for the robostness of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.

Mediation based network approaches are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction. But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error. Recent advances in MR methodology, for example network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust solutions alternatives for these more complicated problems.

There are a number of limitations to the analysis that we performed. We focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present. To give an example in which this might arise consider the case that C-reactive protein (CRP) is being tested for causal association with some outcome, and the FTO variant is used as an instrument. Though FTO is robustly associated with CRP, it is through the upstream influence of BMI [@Timpson2011]. Thus, if BMI has a direct influence on the outcome, the conditions of MR will be violated and the Steiger test proposed here will be unreliable.

### 2016-02-25 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Determining the direction of causality in the face of measurement error},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\setlength{\droptitle}{-2em}
  \title{Determining the direction of causality in the face of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{25 February 2016}

@@

Here, however, arises a potential issue: often the biological effect of
a SNP is not known. If there exists a putative association between two
variables, with a SNP being robustly associated with each, it can be
difficult to determine which of the two variables is subject to the
direct effect of the SNP (i.e., for which of the two variables is the
SNP a valid instrument?). By definition, we expect that if the
association is causal then a SNP for the exposure will be associated
with the outcome, so if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. For example genome wide association studies
(GWASs) that identify genetic associations for complex traits are, by
design, hypothesis free and agnostic of genomic function, and it often
takes years of follow up studies to understand the biological nature of
a putative GWAS hit\textsuperscript{24}. Another situation is where the
causal direction between 'omic measures need to be determined, for
example if a DNA methylation probe is associated with expression of an
adjacent gene, then is a cis-acting SNP an instrument for the DNA
methylation level, or the gene expression level\textsuperscript{8}?

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{25} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{1} is implemented in the R package
\emph{R/cit}\textsuperscript{26}. The \emph{cit.cp} function was used to
obtain an omnibus p-value of causality between an exposure \(x\) and an
outcome \(y\) using a mediator \(g\) by making inferences based on a
number of conditions that we examine in the results. To infer the
direction of causality using the CIT method, an omnibus p-value
generated by CIT, \(p_{CIT}\), was estimated for the correct model of
\(x\) causing \(y\), and for incorrect model of \(y\) causing \(x\). For
some significance threshold \(\alpha\), causal direction was inferred
based on the following scenarios:

@@

Supposing there are two correlated variables, \(x\) and \(y\), with a
variant \(g\) associated with both. Assuming no horizontal pleiotropy
(i.e. \(g\) only has an effect on one of the variables through the other
variable) it is desirable to know which of the variables it has a direct
influence on. This can be achieved by testing for a difference in the
correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
for correlated correlations within a population\textsuperscript{27}. It
is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{28} setting, where the Steiger test of two
independent correlations is applied instead of the case detailed here
where correlations share one variable in common. An advantage of using
the Steiger test in the two sample context is that it can compare
correlations in independent samples where sample sizes are different.

@@

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of mediation based methods. Mendelian
randomisation is robust in the face of measurement error and, given that
the researcher has knowledge about the biology of the instrument being
used in the analysis, can offer a direct solution to the issues that CIT
faces. This assumption is often reasonable, for example SNPs are
commonly used as instruments when they are found in genes with known
biological relevance to the trait of interest. But on many occasions
this is not the case, and it may be tempting to resort to using methods
based on mediation in order to be able to both ascertain if there is a
causal association and to infer the direction of causality. Here we have
described a simple extension to MR which can be used as an alternative
to mediation based methods. We show that this method is still liable to
measurement error, but because it has different properties to the CIT it
offers two main advantages. First, it uses a formal statistical
framework to test for the robostness of the assumed direction of
causality. Second, after testing in a comprehensive range of scenarios
the MR based approach is much less likely to infer the wrong direction
of causality compared to CIT.

There are a number of limitations to the analysis that we performed. We
focused on the simple case of a single instrument in a single sample
setting, and assumed that pleiotropy (the influence of the instrument on
the outcome through a mechanism other than the exposure) was not
present.

@@

25.R Core Team. \emph{R: A Language and Environment for Statistical
Computing}. (R Foundation for Statistical Computing, 2015). at
\textless{}\url{https://www.r-project.org/}\textgreater{}

26.Millstein, J. \emph{cit: Causal Inference Test. R package version
1.9}. (2016). at
\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}

27.Steiger, J. H. Tests for comparing elements of a correlation matrix.
\emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).

28.Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
randomization studies: subsample and 2-sample instrumental variable
estimators. \emph{American journal of epidemiology} \textbf{178,}
1177--84 (2013).

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Determining the existance and direction of causality in the face of measurement error},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\setlength{\droptitle}{-2em}
  \title{Determining the existance and direction of causality in the face of
measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{25 February 2016}

@@

Here, however, arises a potential issue: often the biological effect of
a SNP is not known. If there exists a putative association between two
variables, with a SNP being robustly associated with each, it can be
difficult to determine which of the two variables is subject to the
direct effect of the SNP (i.e., for which of the two variables is the
SNP a valid instrument?). By definition, we expect that if the
association is causal then a SNP for the exposure will be associated
with the outcome, so if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. For example genome wide association studies
(GWASs) that identify genetic associations for complex traits are, by
design, hypothesis free and agnostic of genomic function, and it often
takes years of follow up studies to understand the biological nature of
a putative GWAS hit\textsuperscript{24}. To illustrate a trivial example
- the same FTO variant that is famously associated with BMI is also
associated with C-reactive protein (CRP). It is known that this is
because BMI causes CRP\textsuperscript{25}, but supposing we only had
\emph{a priori} knowledge that FTO associated with CRP, and we used it
to instrument CRP in an MR analysis of CRP against BMI, we would infer
that CRP causes BMI. Another situation is where the causal direction
between 'omic measures need to be determined, for example if a DNA
methylation probe is associated with expression of an adjacent gene,
then is a cis-acting SNP an instrument for the DNA methylation level, or
the gene expression level\textsuperscript{8}?

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{26} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{1} is implemented in the R package
\emph{R/cit}\textsuperscript{27}. The \emph{cit.cp} function was used to
obtain an omnibus p-value of causality between an exposure \(x\) and an
outcome \(y\) using a mediator \(g\) by making inferences based on a
number of conditions that we examine in the results. To infer the
direction of causality using the CIT method, an omnibus p-value
generated by CIT, \(p_{CIT}\), was estimated for the correct model of
\(x\) causing \(y\), and for incorrect model of \(y\) causing \(x\). For
some significance threshold \(\alpha\), causal direction was inferred
based on the following scenarios:

@@

Supposing there are two correlated variables, \(x\) and \(y\), with a
variant \(g\) associated with both. Assuming no horizontal pleiotropy
(i.e. \(g\) only has an effect on one of the variables through the other
variable) it is desirable to know which of the variables it has a direct
influence on. This can be achieved by testing for a difference in the
correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
for correlated correlations within a population\textsuperscript{28}. It
is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{29} setting, where the Steiger test of two
independent correlations is applied instead of the case detailed here
where correlations share one variable in common. An advantage of using
the Steiger test in the two sample context is that it can compare
correlations in independent samples where sample sizes are different.

@@

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of mediation based methods. Mendelian
randomisation is robust in the face of measurement error and, given that
the researcher has knowledge about the biology of the instrument being
used in the analysis, can offer a direct solution to the issues that CIT
faces. This assumption is often reasonable, for example SNPs are
commonly used as instruments when they are found in genes with known
biological relevance to the trait of interest. But on many occasions
this is not the case, and it may be tempting to resort to using methods
based on mediation in order to be able to both ascertain if there is a
causal association and to infer the direction of causality. Here we have
described a simple extension to MR which can be used as an alternative
to mediation based methods. We show that this method is still liable to
measurement error, but because it has different properties to the CIT it
offers two main advantages. First, it uses a formal statistical
framework to test for the robostness of the assumed direction of
causality. Second, after testing in a comprehensive range of scenarios
the MR based approach is less likely to infer the wrong direction of
causality compared to CIT, while substantially improving power over CIT
in the cases where \(d > 0\).

Mediation based network approaches are very well
established\textsuperscript{14} and have a number of extensions that
make them valuable tools, including for example network construction.
But because they are predicated on the basic underlying principles of
mediation they are liable to suffer from the same issues of measurement
error. Recent advances in MR methodology, for example network
MR\textsuperscript{30} and mediation through MR\textsuperscript{31} may
offer more robust solutions alternatives for these more complicated
problems.

There are a number of limitations to the analysis that we performed. We
focused on the simple case of a single instrument in a single sample
setting, and assumed that pleiotropy (the influence of the instrument on
the outcome through a mechanism other than the exposure) was not
present. To give an example in which this might arise consider the case
that C-reactive protein (CRP) is being tested for causal association
with some outcome, and the FTO variant is used as an instrument. Though
FTO is robustly associated with CRP, it is through the upstream
influence of BMI\textsuperscript{25}. Thus, if BMI has a direct
influence on the outcome, the conditions of MR will be violated and the
Steiger test proposed here will be unreliable.

@@

25.Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
index: elucidating direction of causation through reciprocal Mendelian
randomization. \emph{International journal of obesity (2005)}
\textbf{35,} 300--8 (2011).

26.R Core Team. \emph{R: A Language and Environment for Statistical
Computing}. (R Foundation for Statistical Computing, 2015). at
\textless{}\url{https://www.r-project.org/}\textgreater{}

27.Millstein, J. \emph{cit: Causal Inference Test. R package version
1.9}. (2016). at
\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}

28.Steiger, J. H. Tests for comparing elements of a correlation matrix.
\emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).

29.Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
randomization studies: subsample and 2-sample instrumental variable
estimators. \emph{American journal of epidemiology} \textbf{178,}
1177--84 (2013).

30.Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
Network Mendelian randomization: using genetic variants as instrumental
variables to investigate mediation in causal pathways.
\emph{International journal of epidemiology} \textbf{44,} 484--95
(2015).

31.Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
cholesterol, and blood pressure as mediators from obesity to ischemic
heart disease. \emph{Circulation research} \textbf{116,} 665--73 (2015).
